### 2022-10-24 response.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\ensuremath{{\mathcal R}_{0}}\xspace}
\newcommand{\Rs}{\Rx{\mathrm{s}}}
\newcommand{\Rpool}{\ensuremath{{\mathcal R}_{\textrm{\tiny{pool}}}}\xspace}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript.

\revtext{Editor review}
Dear Mr. Park,

Thank you for submitting your manuscript Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants of SARS-CoV-2 for consideration by PNAS. We apologize for the delay in assessing your manuscript. After an evaluation of its suitability for PNAS, your manuscript was assigned to an editor with expertise in the subject area of your work, who examined the manuscript and identified 2 expert reviewers in the field to assess your work. These reviews, which appear below, were considered by the editor, who determined that the issues raised by the reviewers were of sufficient concern to merit rejection of the manuscript.

The reviewers were positive, however, about certain aspects of the work, an assessment with which the editor concurs; accordingly, we are willing to consider a resubmission that constructively addresses all of the concerns raised in the reviews. The revised manuscript must satisfy both the reviewers and the editor in order to be accepted. Please know, though, that we cannot guarantee acceptance because new criticisms could arise upon re-evaluation, and another round of resubmission is not allowed.

If you would like to submit a resubmission, please submit it as a new paper. In your cover letter and the Comments to Editorial Staff, please specify that the paper is a resubmission and include your original tracking number and a point-by-point response to the editor and reviewers' comments. Your resubmission will be given a new tracking number. If you have any questions, please contact our office.

Dear Colleagues:

Both reviewers, as well as I, like this report and the way it is written. Reviewer 2 recommended its acceptance. Reviewer 1 raised an issue you need to address, "WHY the omicron variant had such a short generation time in this setting...", and 
recommend
 the rejection of your report. I agree with this review about this issue.

Based on that reviewer's criticism, I am recommending what I call a soft rejection. Address this reviewer's criticism in detail and modify the manuscript accordingly. By "address," I am considering the possibility of 
your
 presenting a compelling argument for why this issue need not be considered in this report. Your response to the reviews and the revised manuscript will be sent to the second reviewer.

I look forward to reading your revised manuscript.

Best wishes for your well-being and that of all the people you care for whom you care,

Be well,

Bruce

\rev{Reviewer \#1}

\revtext{The paper commences with a very clear introduction to the problems in estimating the transmission advantage of a new variant - in particular, the role of dynamical bias. This is followed by a very reasonable proposal for how the problem may be addressed. Applying these methods to a re-analysis of data on within- and between- household spread of delta and omicron variants of SARS-CoV-2 suggests that while both had similar incubation-period distributions, the mean generation interval for omicron was shorter by 0.3-0.8 days than previously estimated (0.6 days). The original analysis had also reported a difference in mean incubation periods of 3.2 days for omicron versus 4.4 days for delta; the methods used in this paper suggest that both have a mean of 4.1 days. The consequences of discriminating between generation time intervals on the estimation of transmission advantage is shown in Fig 5, although it is not clear how this pertains to dynamical bias as both the original study and this one reported a difference between the variants in this respect.}

Dynamical biases pertain to the estimation of delay distributions

\revtext{The caveats of this study, and the problems of applying this method to other datasets, are very thoroughly outlined in the discussion.}

Thank you.

\revtext{Less attention is paid to the more important question of WHY the omicron variant had such a short generation time in this setting.}

\revtext{The authors discuss the possibility of a "network effect" which is predicated on omicron having a higher Ro and therefore somewhat circular.}

Not circular.

\revtext{Other factors "such as more stringent contact tracing measures against the Omicron variant in the Netherlands [4], faster within-host clearance of the Omicron variant [35], and viral kinetics of reinfection and breakthrough infections" are mentioned without sufficient explanation as to why they would yield such a result.}

Can provide more detail.

\revtext{The role of immune evasion - although mentioned in passing - has not been sufficiently accommodated either in the discussion or in the analysis itself, even though it presents itself as the most likely explanation for the success of omicron in almost every global setting.}

Yes and no.

\revtext{To address this properly, the authors will have to account for the different proportions of the population currently immune to infection by either variant and how the level of cross-immunity between them affects our observations of transmission advantage. The authors should at least justify why they have not taken such an obvious feature of this multi-strain system into account.}

\rev{Reviewer \#2}

\revtext{The significance statement is accurate. I find the paper definitely relevant for future epidemiological studies on COVID-19, and even important for some technical works. But neither the methods nor the results are likely to have a huge impact on the field, even less on other fields.}

\revtext{The manuscript is technically sound and very well written. The results are convincing and clearly explained in the manuscript, and the arguments presented make sense from an epidemiological perspective. This paper emphasizes once more the importance of accounting for the epidemic dynamic (or the differential dynamic between two variants, as in this case) when inferring fundamental epidemiologically relevant quantities that vary in time during the infection. I find it an important contribution to the literature, albeit a technical one.}

\end{document}

### 2022-11-02 response.tex

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript.

\rev{Editor}

\revtext{Dear Colleagues:}

\revtext{Both reviewers, as well as I, like this report and the way it is written. Reviewer 2 recommended its acceptance. Reviewer 1 raised an issue you need to address, "WHY the omicron variant had such a short generation time in this setting...", and 
recommend the rejection of your report. I agree with this review about this issue.}

\revtext{Based on that reviewer's criticism, I am recommending what I call a soft rejection. Address this reviewer's criticism in detail and modify the manuscript accordingly. By "address," I am considering the possibility of your presenting a compelling argument for why this issue need not be considered in this report. Your response to the reviews and the revised manuscript will be sent to the second reviewer.}

### 2022-11-02 omicron-generation.tex
ack
done editing?

A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections. 
Previous simulations showed that network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
Other factors, such as more stringent contact tracing measures against the Omicron variant in the Netherlands \citep{backer2021omicron} faster within-host clearance of the Omicron variant \citep{hay2022viral}, and viral kinetics of reinfection and breakthrough infections, also likely contributed to shortening of generation intervals.
If shorter generation intervals of the Omicron variant represents an increased proportion of presymptomatic transmission, control measures that target symptomatic individuals can become less effective.

While our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, it is still uncertain how infectiousness profiles differ intrinsically between Omicron and Delta.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
In addition, the ``intrinsic'' generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic (or ``unmitigated'') generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
and can therefore be expected to improve estimates of effective reproduction numbers.

@@

We thank Ron Milo for providing helpful comments on the manuscript. JSW acknowledges support from the Army Research Office W911NF1910384 and the Ile de France region via the Chaires Blaise Pascal program. JW and JB has received funding from European Union's Horizon 2020 research and innovation programme---project EpiPose, Epidemic intelligence to Minimize COVID-19’s Public Health, Societal and Economical Impact (grant agreement number 101003688). S.F. was supported by the Wellcome Trust (grant no. 210758/Z/18/Z). The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript. The findings and conclusions in this study are those of the authors and do not necessarily represent the official position of the funding agencies, the National Institutes of Health, or the U.S. Department of Health and Human Services.

>>>>

Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
The realized generation intervals represent time between actual infection events and are different from the intrinsic generation intervals, which represent the changes in the infectiousness of an infected individuals over the course of their infections \citep{champredon2015intrinsic}.

Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals of the Omicron variant.
For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} can lead to faster recovery, which in turn can also shorten the intrinsic generation interval \citep{roberts2007model}.
Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.

Even if the Delta and Omicron variants were to have identical intrinsic generation-interval distributions, their realized generation-interval distributions can still differ.
For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, can also lead to shorter realized generation intervals.
However, this explanation is also less likely given similarities in the inferred incubation-period distributions.

Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections. 
To better illustrate the network effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
in this case, both variants have identical intrinsic generation-interval distributions.
However, if both individuals were each living in a household consisting of four other people, an individual infected with variant A will infect one household member everyday; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
Previous simulations showed that such network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).

Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but it is still uncertain how infectiousness profiles differ intrinsically between Omicron and Delta.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
and can therefore be expected to improve estimates of effective reproduction numbers.

@@

We thank Ron Milo for providing helpful comments on the manuscript. 
JD was supported by the Canadian Institutes of Health Research, the Natural Sciences and Engineering Research Council of Canada, and the Michael G. DeGroote Institute for Infectious Disease Research.
JSW acknowledges support from the Army Research Office W911NF1910384 and the Ile de France region via the Chaires Blaise Pascal program. 
JW and JB has received funding from European Union's Horizon 2020 research and innovation programme---project EpiPose, Epidemic intelligence to Minimize COVID-19’s Public Health, Societal and Economical Impact (grant agreement number 101003688). 
SF was supported by the Wellcome Trust (grant no. 210758/Z/18/Z). 
The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript. 
The findings and conclusions in this study are those of the authors and do not necessarily represent the official position of the funding agencies, the National Institutes of Health, or the U.S. Department of Health and Human Services.
